---
title: "Simulating Covid-19 Vaccine Data using a Discrete-Time Simulation"
output: rmarkdown::html_vignette
author: "Robin Denz"
vignette: >
  %\VignetteIndexEntry{Simulating Covid-19 Vaccine Data using a Discrete-Time Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>"
)
```

# Introduction


# Data at t0

In our case, the data at the beginning of the simulation contains only the geographic entity `kreis`, the age of the person `sex` and the sex of the person `sex`. The `kreis` and `sex` are treated as root nodes, while the age distribution is generated to be different in each `kreis`, making `age` a child node with `kreis` as a single parent. The `sim_from_dag` function is used to directly generate a dataset using these specifications:

```{r}
library(Rfast)
library(truncnorm)
library(devtools)
library(xlsx)
library(truncnorm)
library(data.table)

load_all()

## node kreis
generate_kreis <- function(n) {
  
  probs_kreis <- rbeta(n=400, shape1=1.4, shape2=0.6)
  probs_kreis <- probs_kreis / sum(probs_kreis)
  
  kreis <- sample(c(1:400), size=n, replace=TRUE, prob=probs_kreis)
  
  return(kreis)
}

## node age
node_generate_age <- function(data, parents, betas=NULL) {
  age <- rep(NA, nrow(data))
  for (i in seq_len(400)) {
    dist_mean <- rtruncnorm(n=1, a=40, b=70, sd=10, mean=55)
    dist_sd <- rtruncnorm(n=1, a=2, b=25, mean=15, sd=10)
    age[data$kreis==i] <- rtruncnorm(n=sum(data$kreis==i),
                                     a=0, b=100,
                                     mean=dist_mean,
                                     sd=dist_sd)
  }
  return(age)
}

t0_root_nodes <- list(list(name="kreis",
                           dist="generate_kreis",
                           params=list()),
                      list(name="sex",
                           dist="rbernoulli",
                           params=list(p=0.5)),
                      list(name="vacc_type",
                           dist="rcategorical",
                           params=list(probs=c(0.73919444, 0.17035600,
                                               0.06936522, 0.02032859),
                                       labels=c(1, 2, 3, 4))))
t0_child_nodes <- list(list(name="age",
                            parents="kreis",
                            type="generate_age"))

start_data <- sim_from_dag(n_sim=10, root_nodes=t0_root_nodes, child_nodes=t0_child_nodes)
```

This is how the resulting data looks:

```{r}
head(start_data)
```

## Data at tx:

Our next task is to define what kind of time-dependent nodes we want to use and how exactly they should work.

```{r}

## node age (to increase it as the simulation goes on)
node_advance_age <- function(data) {
  return(data$age + 1/365)
}

## probability of first vaccination
prob_first_vacc <- function(data, sim_time, start_vaccinations,
                            impf_dat, beta_age) {
  
  if (sim_time < start_vaccinations) {
    return(0)
  }
  # There is only so much data yet, afterwards simply return a small but
  # constant value.
  if (sim_time - start_vaccinations >= 600) {
    return(0.000000001)
  }
  
  p <- impf_dat$impf_prob[impf_dat$time==(sim_time - start_vaccinations + 1)]
  p <- p + data$age * beta_age
  
  return(p)
}

## time of second vaccination
node_generate_vacc_2 <- function(data) {
  vacc_2 <- data$vacc_1_time + ((data$vacc_type <= 2) * 42 + 
                                (data$vacc_type==3) * 84) * 
    NA^(data$vacc_type==4)
  return(vacc_2)
}

## base probability of getting covid 19 depending on time
base_p_covid <- function(sim_time) {
  # where do we get this?
  return(0.001)
}

## probability of getting covid 19, depending on time, age and previous vaccine
prob_covid <- function(data, sim_time, start_covid, rr_vacc_start,
                       dur_vacc, vacc_delay, beta_age) {
  
  if (sim_time < start_covid) {
    return(0)
  }
  
  # Alternatively we could use the linear equation defined by the two points:
  # y = ((y2-y1) / (x2-x1)) * x + ((x2*y1-x1*y2) / (x2-x1))
  # to find the person and time specific RR
  # y2 <- 1
  # y1 <- rr_vacc_start
  x1 <- fifelse(data$vacc_type==4, data$vacc_1_time, data$vacc_2) + vacc_delay
  x2 <- x1 + dur_vacc
  
  vacc_reduction <- rr_vacc_start^pmax((1-(sim_time-x1)/x2)*(sim_time-x1 >=0),0)
  
  # set to 1 if NA
  vacc_reduction[is.na(vacc_reduction)] <- 1
  
  # get probability
  p <- (base_p_covid(sim_time) + data$age * beta_age) * vacc_reduction
  
  return(p)
}

## probability of getting sickness
prob_sickness <- function(data, sim_time, rr_vacc_1, rr_vacc_2,
                          rr_covid, risk_dur_vacc, risk_dur_covid,
                          beta_age, base_p) {
  
  # check if in risk periods
  in_vacc_1_risk <- fifelse(is.na(data$vacc_1_time), FALSE,
                            fifelse((sim_time - data$vacc_1_time) < 
                                    risk_dur_vacc, TRUE, FALSE))
  in_vacc_2_risk <- fifelse(is.na(data$vacc_2), FALSE,
                           fifelse(((sim_time - data$vacc_2) < risk_dur_vacc) & 
                                   ((sim_time - data$vacc_2) > 0), TRUE, FALSE))
  in_covid_risk <- fifelse(is.na(data$covid_time), FALSE,
                           fifelse((sim_time - data$covid_time) < 
                                   risk_dur_covid, TRUE, FALSE))
  
  # choose which relative risk to apply to baseline risk
  time_dep_risk <- fifelse(!in_vacc_1_risk & !in_vacc_2_risk &
                           !in_covid_risk, 1,
                          fifelse(in_covid_risk, rr_covid,
                          fifelse(in_vacc_1_risk, rr_vacc_1[data$vacc_type],
                                  rr_vacc_2[data$vacc_type])))
  
  # put all risks together
  p <- (base_p + data$age * beta_age) * time_dep_risk
    
  return(p)
}

# read and clean Impfdata
impf_dat <- read.xlsx("P:/BMG_Covid19Impfen/Auswertung/Auswertungsplan_Stufen/Simulation_Testdaten/Impfquotenmonitoring.xlsx", sheetIndex=4)
impf_dat <- impf_dat[1:(nrow(impf_dat)-4),]
impf_dat$Datum <- as.Date(impf_dat$Datum, format="%d.%m.%Y")
impf_dat$NA. <- NULL

# probability of first vaccination
impf_dat$time <- 1:nrow(impf_dat)
impf_dat$cumsum <- cumsum(impf_dat$Erstimpfung)
impf_dat$cumsum <- c(0, impf_dat$cumsum[1:(nrow(impf_dat)-1)])

impf_dat$impf_prob <- impf_dat$Erstimpfung / (83020000 - impf_dat$cumsum)

# definition of time-dependent nodes
tx_nodes <- list(list(name="age",
                      parents="age",
                      type="advance_age"),
                 list(name="vacc_1",
                      parents=c("age", "vacc_1_time", "vacc_1_event"),
                      type="time_to_event",
                      prob_fun=prob_first_vacc,
                      prob_fun_args=list(start_vaccinations=365*2,
                                         impf_dat=impf_dat,
                                         beta_age=0.000001),
                      event_duration=Inf,
                      save_past_events=FALSE),
                 list(name="vacc_2",
                      parents=c("vacc_1_event", "vacc_1_time", "vacc_type"),
                      type="generate_vacc_2"),
                 list(name="covid",
                      parents=c("vacc_1_time", "vacc_2", "vacc_type", "age",
                                "covid_event", "covid_time"),
                      type="time_to_event",
                      prob_fun=prob_covid,
                      prob_fun_args=list(start_covid=365,
                                         rr_vacc_start=0.8,
                                         dur_vacc=180,
                                         vacc_delay=14,
                                         beta_age=0.000000001),
                      event_duration=21,
                      immunity_duration=150,
                      save_past_events=TRUE),
                 list(name="sickness",
                      parents=c("vacc_1_event", "vacc_1_time", "vacc_2",
                                "age", "covid_event", "covid_time",
                                "vacc_type", "sickness_event", "sickness_time",
                                "sickness_past_event_times"),
                      type="time_to_event",
                      prob_fun=prob_sickness,
                      prob_fun_args=list(rr_vacc_1=c(3.24, 3.24, 1, 1),
                                         rr_vacc_2=c(3.24, 3.24, 1, 1),
                                         rr_covid=3,
                                         risk_dur_vacc=21,
                                         risk_dur_covid=21,
                                         beta_age=0.00000001,
                                         base_p=0.00000001),
                      event_duration=21,
                      immunity_duration=50,
                      save_past_events=TRUE))

out <- sim_discrete_time(n_sim=100000,
                         t0_root_nodes=t0_root_nodes,
                         t0_child_nodes=t0_child_nodes,
                         t0_sort_dag=TRUE,
                         max_t=365*3,
                         tx_nodes=tx_nodes,
                         save_states="last")$data

```

Using parallel processing to speed things up:

```{r}
library(doParallel)
library(foreach)
library(doRNG)
library(doSNOW)

n_cores <- 8

cl <- parallel::makeCluster(n_cores, outfile="")
doParallel::registerDoParallel(cl)
parallel::clusterExport(cl, ls())
pkgs <- (.packages())

out <- foreach::foreach(i=1:n_cores, .packages=pkgs) %dorng% {
                          
  temp <- sim_discrete_time(n_sim=100000,
                            t0_root_nodes=t0_root_nodes,
                            t0_child_nodes=t0_child_nodes,
                            t0_sort_dag=TRUE,
                            max_t=365*3,
                            tx_nodes=tx_nodes,
                            save_states="last")$data
}
parallel::stopCluster(cl)
```


