---
title: "Simulating Covid-19 Vaccine Data using a Discrete-Time Simulation"
output: rmarkdown::html_vignette
author: "Robin Denz"
vignette: >
  %\VignetteIndexEntry{Simulating Covid-19 Vaccine Data using a Discrete-Time Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>"
)
```

# Introduction

# How it generally works

All variables in a discrete-time simulation can be categorized into three categories: *t0_root nodes*, *t0_child_nodes* and *tx_nodes*.

* *t0_root_nodes*: Variables that are completely independent of all other variables and are only generated once are called *t0_root_nodes*. This could be something like sex or geographic entities. They are generally simply sampled from some previously defined distribution, but they could also be sampled directly from existing data. The prefix *t0_* indicates that these variables are created only once in the beginning.

* *t0_child_nodes*: Much like *t0_root_nodes*, the *t0_child_nodes* are also variables that are generated only once in the beginning of the discrete-time simulation. The only difference is, that these variables are not simply sampled from a defined distribution. Instead they causally depend in some way on other variables. Those other variables can be root nodes or other child nodes, as long as the underlying causal structure can be described as a directed acyclic graph.

* *tx_nodes*: This type of node is more complex and is the reason to use discrete-time simulations. Variables in this category are updated at each step in time of the simulation process. These updates can be fairly easy, such as increasing the age of a person by one time unit on each step, or they can be very complex. For example, if we want to model the occurrence of some kind of time-dependent-event, we can generate the probability of that occurrence at each step depending on other variables in the simulation model or past states of the variable itself.

As a natural consequence, the simulation model follows the following steps:

* **1**: Sampling from the *t0_root_nodes*
* **2**: Generating the *t0_child_nodes* based on root nodes and other child nodes
* **3**: Increasing the simulation time by 1
* **4**: Generating / updating *tx_nodes*
* **5**: Repeating steps 3 and 4 until a previously defined maximum time point is reached.

# Simple Example

## Data at t0

## Data at tx


# Complex Example

## Data at t0

In our case, the data at the beginning of the simulation contains only the geographic entity `kreis`, the age of the person `sex` and the sex of the person `sex`. The `kreis` and `sex` are treated as root nodes, while the age distribution is generated to be different in each `kreis`, making `age` a child node with `kreis` as a single parent. The `sim_from_dag` function is used to directly generate a dataset using these specifications:

```{r}
library(Rfast)
library(truncnorm)
library(devtools)
library(xlsx)
library(truncnorm)
library(data.table)

load_all()

## node kreis
generate_kreis <- function(n) {
  
  probs_kreis <- rbeta(n=400, shape1=1.4, shape2=0.6)
  probs_kreis <- probs_kreis / sum(probs_kreis)
  
  kreis <- sample(c(1:400), size=n, replace=TRUE, prob=probs_kreis)
  
  return(kreis)
}

## node age
node_generate_age <- function(data, parents, betas=NULL) {
  age <- rep(NA, nrow(data))
  for (i in seq_len(400)) {
    dist_mean <- rtruncnorm(n=1, a=40, b=70, sd=10, mean=55)
    dist_sd <- rtruncnorm(n=1, a=2, b=25, mean=15, sd=10)
    age[data$kreis==i] <- rtruncnorm(n=sum(data$kreis==i),
                                     a=0, b=100,
                                     mean=dist_mean,
                                     sd=dist_sd)
  }
  return(age)
}

t0_root_nodes <- list(list(name="kreis",
                           dist="generate_kreis",
                           params=list()),
                      list(name="sex",
                           dist="rbernoulli",
                           params=list(p=0.5)),
                      list(name="vacc_type",
                           dist="rcategorical",
                           params=list(probs=c(0.73919444, 0.17035600,
                                               0.06936522, 0.02032859),
                                       labels=c(1, 2, 3, 4))))
t0_child_nodes <- list(list(name="age",
                            parents="kreis",
                            type="generate_age"))

start_data <- sim_from_dag(n_sim=10, root_nodes=t0_root_nodes, child_nodes=t0_child_nodes)
```

This is how the resulting data looks:

```{r}
head(start_data)
```

## Data at tx:

Our next task is to define what kind of time-dependent nodes we want to use and how exactly they should work.

```{r}

## node age (to increase it as the simulation goes on)
node_advance_age <- function(data) {
  return(data$age + 1/365)
}

## probability of first vaccination
prob_first_vacc <- function(data, sim_time, start_vaccinations,
                            impf_dat, beta_age) {
  
  if (sim_time < start_vaccinations) {
    return(0)
  }
  # There is only so much data yet, afterwards simply return a small but
  # constant value.
  if (sim_time - start_vaccinations >= 600) {
    return(0.000000001)
  }
  
  p <- impf_dat$impf_prob[impf_dat$time==(sim_time - start_vaccinations + 1)]
  p <- p + data$age * beta_age
  
  return(p)
}

## time of second vaccination
node_generate_vacc_2 <- function(data) {
  vacc_2 <- data$vacc_1_time + ((data$vacc_type <= 2) * 42 + 
                                (data$vacc_type==3) * 84) * 
    NA^(data$vacc_type==4)
  return(vacc_2)
}

## base probability of getting covid 19 depending on time
base_p_covid <- function(sim_time) {
  # where do we get this?
  return(0.001)
}

## probability of getting covid 19, depending on time, age and previous vaccine
prob_covid <- function(data, sim_time, start_covid, rr_vacc_start,
                       dur_vacc, vacc_delay, beta_age) {
  
  if (sim_time < start_covid) {
    return(0)
  }
  
  # Alternatively we could use the linear equation defined by the two points:
  # y = ((y2-y1) / (x2-x1)) * x + ((x2*y1-x1*y2) / (x2-x1))
  # to find the person and time specific RR
  # y2 <- 1
  # y1 <- rr_vacc_start
  x1 <- fifelse(data$vacc_type==4, data$vacc_1_time, data$vacc_2) + vacc_delay
  x2 <- x1 + dur_vacc
  
  vacc_reduction <- rr_vacc_start^pmax((1-(sim_time-x1)/x2)*(sim_time-x1 >=0),0)
  
  # set to 1 if NA
  vacc_reduction[is.na(vacc_reduction)] <- 1
  
  # get probability
  p <- (base_p_covid(sim_time) + data$age * beta_age) * vacc_reduction
  
  return(p)
}

## probability of getting sickness
prob_sickness <- function(data, sim_time, rr_vacc_1, rr_vacc_2,
                          rr_covid, risk_dur_vacc, risk_dur_covid,
                          beta_age, base_p) {
  
  # check if in risk periods
  in_vacc_1_risk <- fifelse(is.na(data$vacc_1_time), FALSE,
                            fifelse((sim_time - data$vacc_1_time) < 
                                    risk_dur_vacc, TRUE, FALSE))
  in_vacc_2_risk <- fifelse(is.na(data$vacc_2), FALSE,
                           fifelse(((sim_time - data$vacc_2) < risk_dur_vacc) & 
                                   ((sim_time - data$vacc_2) > 0), TRUE, FALSE))
  in_covid_risk <- fifelse(is.na(data$covid_time), FALSE,
                           fifelse((sim_time - data$covid_time) < 
                                   risk_dur_covid, TRUE, FALSE))
  
  # choose which relative risk to apply to baseline risk
  time_dep_risk <- fifelse(!in_vacc_1_risk & !in_vacc_2_risk &
                           !in_covid_risk, 1,
                          fifelse(in_covid_risk, rr_covid,
                          fifelse(in_vacc_1_risk, rr_vacc_1[data$vacc_type],
                                  rr_vacc_2[data$vacc_type])))
  
  # put all risks together
  p <- (base_p + data$age * beta_age) * time_dep_risk
    
  return(p)
}

# read and clean Impfdata
impf_dat <- read.xlsx("P:/BMG_Covid19Impfen/Auswertung/Auswertungsplan_Stufen/Simulation_Testdaten/Impfquotenmonitoring.xlsx", sheetIndex=4)
impf_dat <- impf_dat[1:(nrow(impf_dat)-4),]
impf_dat$Datum <- as.Date(impf_dat$Datum, format="%d.%m.%Y")
impf_dat$NA. <- NULL

# probability of first vaccination
impf_dat$time <- 1:nrow(impf_dat)
impf_dat$cumsum <- cumsum(impf_dat$Erstimpfung)
impf_dat$cumsum <- c(0, impf_dat$cumsum[1:(nrow(impf_dat)-1)])

impf_dat$impf_prob <- impf_dat$Erstimpfung / (83020000 - impf_dat$cumsum)

# definition of time-dependent nodes
tx_nodes <- list(list(name="age",
                      parents="age",
                      type="advance_age"),
                 list(name="vacc_1",
                      parents=c("age", "vacc_1_time", "vacc_1_event"),
                      type="time_to_event",
                      prob_fun=prob_first_vacc,
                      prob_fun_args=list(start_vaccinations=365*2,
                                         impf_dat=impf_dat,
                                         beta_age=0.000001),
                      event_duration=Inf,
                      save_past_events=FALSE),
                 list(name="vacc_2",
                      parents=c("vacc_1_event", "vacc_1_time", "vacc_type"),
                      type="generate_vacc_2"),
                 list(name="covid",
                      parents=c("vacc_1_time", "vacc_2", "vacc_type", "age",
                                "covid_event", "covid_time"),
                      type="time_to_event",
                      prob_fun=prob_covid,
                      prob_fun_args=list(start_covid=365,
                                         rr_vacc_start=0.8,
                                         dur_vacc=180,
                                         vacc_delay=14,
                                         beta_age=0.000000001),
                      event_duration=21,
                      immunity_duration=150,
                      save_past_events=TRUE),
                 list(name="sickness",
                      parents=c("vacc_1_event", "vacc_1_time", "vacc_2",
                                "age", "covid_event", "covid_time",
                                "vacc_type", "sickness_event", "sickness_time",
                                "sickness_past_event_times"),
                      type="time_to_event",
                      prob_fun=prob_sickness,
                      prob_fun_args=list(rr_vacc_1=c(3.24, 3.24, 1, 1),
                                         rr_vacc_2=c(3.24, 3.24, 1, 1),
                                         rr_covid=3,
                                         risk_dur_vacc=21,
                                         risk_dur_covid=21,
                                         beta_age=0.00000001,
                                         base_p=0.00000001),
                      event_duration=21,
                      immunity_duration=50,
                      save_past_events=TRUE))

out <- sim_discrete_time(n_sim=100000,
                         t0_root_nodes=t0_root_nodes,
                         t0_child_nodes=t0_child_nodes,
                         t0_sort_dag=TRUE,
                         max_t=365*3,
                         tx_nodes=tx_nodes,
                         save_states="last")$data

```

Using parallel processing to speed things up:

```{r}
library(doParallel)
library(foreach)
library(doRNG)
library(doSNOW)

n_cores <- 8

cl <- parallel::makeCluster(n_cores, outfile="")
doParallel::registerDoParallel(cl)
parallel::clusterExport(cl, ls())
pkgs <- (.packages())

out <- foreach::foreach(i=1:n_cores, .packages=pkgs) %dorng% {
                          
  temp <- sim_discrete_time(n_sim=100000,
                            t0_root_nodes=t0_root_nodes,
                            t0_child_nodes=t0_child_nodes,
                            t0_sort_dag=TRUE,
                            max_t=365*3,
                            tx_nodes=tx_nodes,
                            save_states="last")$data
}
parallel::stopCluster(cl)
```


